Pesticide exposure in small streams in Germany

### 2016-09-14 article.tex
More discussion & results...

%% -------------------------------------------------------------------------
\section{Introduction}
\todo{Check all references to supplement}.
More than 50\% of the total land area in Germany are used by agriculture \citep{statistisches_bundesamt_bodenflache_2014}.
In the year 2014 more than 45,000 tonnes of 766 authorized pesticides were sold for application on this area \citep{bundesamt_fur_verbraucherschutz_und_lebensmittelsicherheit_absatz_2015}.
The applied pesticides may enter surface waters via spray-drift, edge-off-field run-off or drainage \citep{stehle_probabilistic_2013,schulz_comparison_2001,liess_determination_1999}.
Especially run-off after heavy precipitation events has been shown to be one of the major input routes for pesticides \citep{schulz_field_2004}.
Once entered the surface waters they may have adverse effects on biota and ecosystem functioning \citep{schafer_thresholds_2012}. 
Although, it is known that pesticides pollution and its effect increase with agricultural land-use \todo{citation needed}, studies investigating the shape of this relationships are missing.

@@

Small waters (SW) comprise a major fraction of streams \citep{nadeau_hydrological_2007}, accommodate a higher proportion of biodiversity compared to larger freshwater systems \citep{davies_comparison_2008, biggs_report_2014} and play an important role in recolonization of disturbed downstream reaches \citep{liess_analyzing_2005, orlinskiy_forested_2015}.
However, SWB might be also at high risk of pesticide contamination from adjacent agricultural areas and lower dilution potential \citep{schulz_field_2004}.
It has been shown that SW are more polluted than bigger streams \citep{stehle_pesticide_2015,schulz_field_2004}.
Despite their relevance only a small fraction of studies were conducted on pesticide pollution of SW, with only few large scale studies \citep{lorenz_specifics_2016}.

In this study we try to fill this gap and analyze large scale nation-wide chemical monitoring data from Germany.
First, we revise the available monitoring data if it is suitable for a large scale description of pesticide pollution.
Then we analyse the relationship between agricultural land use and catchment size with pesticide pollution and investigate potential thresholds.
We also investigate whether samples taken during or shortly after heavy rainfall events or different seasons are more likely to indicate chemical pollution.
Finally, we give an overview on the pollution of SWB in Germany.

@@

\subsection{Effect of precipitation on pesticide exposure}
The spatio-temporal intersection revealed that 5\% of the samples were taken at or after days with rainfall events greater than 10mm / day (Supplement, Figure S3).
Precipitation at the day before ($precip_{-1}$) sampling increased $\mu$ and also $\pi$ for nearly all modeled compounds. 
The effects were less pronounced ($\pi$) or not directed ($\mu$) for precipitation at the day of sampling ($precip_{0}$)(Figure \ref{fig:fig5}, top).

Quarters 2-4 showed generally higher $\mu$ than in Q1. 
Exceptions were Diflufenican, that showed lower RQ in Q2-Q4 compared to Q1.
Smaller negative effects were found for Clothianidin (Q2) and Isoproturon (Q3) (Figure \ref{fig:fig5}, bottom and Supplement Table S4).
We found similar effects on $\pi$.

@@

\subsection{Pesticides in small water bodies}
The dataset comprised 12,887 samples from 1313 small water bodies.
1173 samples (0.3\% of all and 5.6\% of samples with detects) RQ higher then 1 were observed.
Neonicotinoid Insecticides and Chlorpyrifos showed highest risk quotients (Figure \ref{fig:fig6}).
For Thiacloprid, Imidacloprid and Chlorpyrifos RAC was less than LOQ, therefore, all detections have a RQ \textgreater 1. 
The herbicides Nicosulfuron and Diflufenican, as well as the fungicide Dimoxystrobin showed also high exceedances of RQ (29.5/14.2/21.4 \% of all samples with detects).
Highest RQ were observed for Chlorpyrifos (max(RQ) = 244), Dimoxystrobin(max(RQ) = 117) and Isoproturon (max(RQ) = 80). 
Metabolites were most commonly detected in samples (Metazachlor sulfonix acid was in 82\% of all samples were it was analysed) (Supplement Figure S8)
Glyphosate was the compound with highest detection rates, followed by Boscalid and Isoproturon. 
However, only the latter showed RQ exceedances (Figure \ref{fig:fig6}).
In 44.8\% of samples more then one compound was quantified, with a maximum of 54 different compounds in one samples (supplemental figure S7).

@@

@@

%%% Word count
% abstract    :
% 3 big(600)  : 1800
% 2 small(300): 600
% text body   : 3000 
% ackno       : 24
% ======================================
% total       :

>>>>

%% -------------------------------------------------------------------------
\section{Introduction}
\todo{Check all references to supplement}.
More than 50\% of the total land area in Germany are used by agriculture \citep{statistisches_bundesamt_bodenflache_2014}.
In the year 2014 more than 45,000 tonnes of 766 authorized pesticides were sold for application on this area \citep{bundesamt_fur_verbraucherschutz_und_lebensmittelsicherheit_absatz_2015}.
The applied pesticides may enter surface waters via spray-drift, edge-off-field run-off or drainage \citep{stehle_probabilistic_2013,schulz_comparison_2001,liess_determination_1999}.
Especially run-off after heavy precipitation events has been shown to be one of the major input routes for pesticides \citep{schulz_field_2004}.
Once entered the surface waters they may have adverse effects on biota and ecosystem functioning \citep{schafer_thresholds_2012}. 
Although, it is known that pesticides pollution and its effect increase with agricultural land-use \citep{schulz_field_2004}, studies investigating the shape of this relationships are missing.

@@

Small water bodies (SWB) comprise a major fraction of streams \citep{nadeau_hydrological_2007}, accommodate a higher proportion of biodiversity compared to larger freshwater systems \citep{davies_comparison_2008, biggs_report_2014} and play an important role in recolonization of disturbed downstream reaches \citep{liess_analyzing_2005, orlinskiy_forested_2015}.
However, SWB might be also at high risk of pesticide contamination from adjacent agricultural areas and lower dilution potential \citep{schulz_field_2004,liess_determination_1999}.
It has been shown that SWB are more polluted than bigger streams \citep{stehle_pesticide_2015,schulz_field_2004}.
Especially, compounds from agricultural use show seasonal rain-event-driven peak concentrations in SWB \citep{wittmer_significance_2010}.
Despite their biological relevance and potential pesticide exposure only a small fraction of studies were conducted on pesticide pollution of SWB, with only few large scale studies \citep{lorenz_specifics_2016}. 
Moreover, it is currently unknown how precipitation might influence measured pesticide concentrations in national monitoring programs.

In this study we try to fill these gaps and analyze large scale nation-wide chemical monitoring data from Germany.
First, we revise the available monitoring data if it is suitable for a large scale description of pesticide pollution.
Then we analyse the relationship between agricultural land use and catchment size with pesticide pollution and investigate potential thresholds.
We also investigate whether samples taken during or shortly after heavy rainfall events or different seasons are more likely to indicate pesticide pollution.
Finally, we give an overview on pesticide pollution of SWB in Germany.

@@

\subsection{Effect of precipitation on pesticide exposure}
The spatio-temporal intersection revealed that 5\% of the samples were taken at or after days with rainfall events greater than 10mm / day (Supplement, Figure S3).
Precipitation at the day before ($precip_{-1}$) sampling increased $\mu$ and also $\pi$ for nearly all analysed compounds. 
The effects were less pronounced ($\pi$) or not directed ($\mu$) for precipitation at the day of sampling ($precip_{0}$)(Figure \ref{fig:fig5}, top).

Quarters 2-4 showed generally higher $\mu$ than in Q1. 
Exceptions were Diflufenican, that showed significantly lower RQ in Q2-Q4 compared to Q1.
Smaller negative effects were found for Clothianidin (Q2) and Isoproturon (Q3) (Figure \ref{fig:fig5}, bottom and Supplement Table S4).
We found similar effects on $\pi$.

@@

\subsection{Pesticides in small water bodies}
The dataset comprised 12,710 samples from 1295 small water bodies.
1173 samples (0.3\% of all and 5.6\% of samples with detects) RQ higher then 1 were observed.
Neonicotinoid Insecticides and Chlorpyrifos showed highest risk quotients (Figure \ref{fig:fig6}).
For Thiacloprid, Imidacloprid and Chlorpyrifos RAC was less than LOQ, therefore, all detections have a RQ \textgreater 1. 
The herbicides Nicosulfuron and Diflufenican, as well as the fungicide Dimoxystrobin showed also high exceedances of RQ (29.5/14.2/21.4 \% of all samples with detects).
Highest RQ were observed for Chlorpyrifos (max(RQ) = 244), Dimoxystrobin(max(RQ) = 117) and Isoproturon (max(RQ) = 80). 
Metabolites were most commonly detected in samples (Metazachlor sulfonix acid was in 82\% of all samples were it was analysed) (Supplement Figure S8)
Glyphosate was the compound with highest detection rates, followed by Boscalid and Isoproturon. 
However, only the latter showed RQ exceedances (Figure \ref{fig:fig6}).
In 44.8\% of samples more then one compound was quantified, with a maximum of 54 different compounds in one samples (supplemental figure S7).

@@

% current problems in monitoring and possible solutions
Nevertheless, a nationwide assessment of pesticide pollution is hampered by the inhomogeneity of monitoring data between federal states:
There are not only big differences in the spatial distribution and quantity of sampling sites (Figure \ref{fig:fig1}), but also the spectrum of analyzed compounds (Figure \ref{fig:fig2}) and differences in the quality of chemical analyses. 
For Thiacloprid, Imidaclorpid and Chlorpyrifos LOQ were above RAC.
For these compounds a lowering of LOQ is essential for reliable assessment.
Moreover, would a nation-wide assessment benefit from a harmonized spectrum of analysed compounds.

\todo{unfertig}

\subsection{Thresholds for agricultural land use and catchment size}
% agriculture
We found a strong influence of agriculture on the pollution of streams.
If there is more the 25\% agriculture within a catchment pesticides it is likely that RAC will be exceeded with an increase in fully agricultural catchments (above 75 \% agriculture).
To our knowledge this is the first study investigating such thresholds of pesticide exposure.
Thresholds for agricultural land use have only been investigated in the past for biological communities.
\citet{feld_response_2013} found change points of biological community metric for agricultural land use at 40\% in low lands. 
Similarly, \citet{waite_agricultural_2014} found a threshold for aquatic diatoms at 40\%.
Our results coincide with these thresholds and suggest that pesticides might contribute to the observed biological effects.

% size
We did not find a relationship between pesticide pollution and catchment size.
However, previous studies have shown that SWB are more polluted that bigger streams \citep{schulz_field_2004,stehle_pesticide_2015,knauer_pesticides_2016}.
This could be explained by the relatively short gradient of catchment sizes in our dataset, with most of the streams being \textless $100~km^2$ (Figure \ref{fig:fig3}, top).
For example the gradient of \citet{schulz_field_2004} covered 6 orders of magnitude.
Another reason might be the the unequal distribution of catchment sizes, with less sites \textless $10~km^2$ and \textgreater $100~km^2$ (Figure \ref{fig:fig3}, top).

\subsection{Effect of precipitation on pesticide exposure}
Our results revealed that pesticide sampling for chemical monitoring in Germany is mainly performed when no precipitation occurs. 
However, we found higher pesticide exposure if samples where taken at the after heavy rainfall events. 
Interestingly, samples take at they day of a heavy-rainfall-event did not show such a strong relationship.
This could be explained by a sampling just before the actual rainfall event.
Pesticide concentration in agricultural SWB show short term peak concentrations \citep{wittmer_significance_2010}.
Overall, our results indicate that current chemical monitoring strongly underestimates pesticide exposure.

\todo{add alternatives - passive sampler, automatic samplers...}

We found pesticide specific patterns in seasonality of the studied compounds.
Most pesticides showed an increase from April - September, which coincides with their main application season.
Hence, we also found the hightest RQ for the herbicide Diflufenican during the winter/spring season, which is application period for which it is registered \todo{cite BVL}.

\subsection{Pesticides in small water bodies}
Our results suggest that SWB are frequently exposed to biologically relevant pesticide concentrations.
\citet{stehle_pesticide_2015} found the highest percentage of RAC exceedances for organophosphate insecticides. 
By contrast, our results revealed that neonicotinoid insecticides show high exceedances, followed by the organophosphate chlorpyrifos. 
This difference can be attributed to the low sample size for neonicotinoid insecticides in their study (n = 33) compared to the dataset presented here (between 1290 and 2113 samples in SWB, Figure \ref{fig:fig6}). 
Our results shows that this particular class of insecticides may currently pose a high risk to freshwater ecosystems.
Compared to \citet{stehle_pesticide_2015} we found much lower rates RAC exceedances (0.3\% of 372,304 samples with RAC vs 44\% of 1,566 samples). 
This can be attributed to different aims of the data sources: scientific research aims at finding pollutants, whereas monitoring aims mainly at surveillance of water quality, also during periods with lower pesticide usage and at natural sites. 
Contrary, \citet{knauer_pesticides_2016} found exceedances from monitoring data mainly for herbicides and fungicides and only one insecticide Chlorpyrifos.
This might reflect differences in pesticide use between countries and defined RACs.

From the definition of RAC it follows that if the concentration of a compound exceeds its RAC ecological effects are expected.
Accordingly, \citet{stehle_agricultural_2015} found that biological diversity is significantly reduced at RQ of 1.12.
We found RQ values greater than 1.12 at 25\% of all SWB sites. 
Consequently, we conclude that pesticides from agricultural land use are a major treat to small water bodies, the biodiversity they host and services they provide. 
Additional treat arises when taking into account that most pesticides do not occur individually but in mixtures \cite{schreiner_pesticide_2016} and may co-occur with other stressors \citep{schafer_contribution_2016}.

% Approval / Risk Assessment
Monitoring data, despite the outlined limitations, provides an opportunity to study environmental occurrences of pesticides after approval.
The high exceedances of RAC indicate that the approval process for pesticides must be checked and refined.

@@

%%% Word count
% abstract    :
% 4 big(600)  : 2400
% 2 small(300): 600
% text body   : 3000 
% ackno       : 24
% ======================================
% total       :
